Lazy Linearity for a Core Functional Language

### 2025-10-16 popl26.tex
revised case rule typesetting

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D}\s_1\dots x_n{:}_{\D}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for typing semantic linearity in the presence of laziness that we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default/wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting which pattern-bound variables must be
consumed linearly or unrestrictedly.
Additionally, the set of patterns in a case expression is assumed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

@@

\vspace{-0.5cm}
\[
\TypeVarDelta
\]
The rule states that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is made up exactly of the resources in the usage environment of
that variable.
%
\ROUNDTWO{$\Delta$-bound variables} are unrestricted and can be discarded and duplicated, despite
multiple occurrences of the \ROUNDTWO{variable} in an expression not
being well-typed
\ROUNDTWO{due to} non-linear usage of linear resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

@@

In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
resources used in let-bound expressions are only consumed when the respective
binders are evaluated in the let body.
%, i.e. linear resources required by let-bound expressions are consumed lazily.
%
Moreover, resources captured by a let-bound expression cannot be used
simultaneously with the let-binder in its body, since said
resources would end up being consumed more than once and violate (semantic)
linearity. The binder has to be used in mutual exclusion with the linear
resources captured by the expression it binds -- either the binding or the
resources \emph{must} be used.

% Indeed, usage environments allow us to encode mutual exclusivity between
% alternative ways of consuming linear resources (between $\D$-vars and direct
% resource usage). 
Let-bound variables are the canonical example of a
$\Delta$-bound variable.
% , that is, let-bound variables bind expressions in which
% the resources required to type them are consumed lazily rather than eagerly.
%
Annotating let-bound variables with a usage environment $\D$ delays the
consumption of resources to when the variables themselves are used and
guarantees the desired mutual exclusion property between the binder and the resources:
\[
\TypeLet
\]
The rule for (non-recursive) let bindings splits the linear environment in $\D$
and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
surprisingly, the resources $\D$ used to type $e$ are still available in the
environment to type the let body $e'$, alongside the unrestricted $x$ binding
annotated with the usage environment $\D$. In essence, the resources of $e$ being
available in $e'$ reflects how typing the body of a lazy binder
doesn't consume resources, and the binding $x$ being an unrestricted
$\D$-variable ensures that using it will consume the resources $\D$ the
expression captures, \emph{if} it is used.
% \todo[inline]{Let bindings are hard, if they are used then we use resources. If
% they don't get used then we use no resources! In practice, resources that show
% up in the body of the let must be used, be it by using the let binder, or by
% using them directly. This makes the let binder and the resources in its body
% mutual exclusive.}

>>>>

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D}\s_1\dots x_n{:}_{\D}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \Rightarrow e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for typing semantic linearity in the presence of laziness that we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default/wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting which pattern-bound variables must be
consumed linearly or unrestrictedly.
Additionally, the set of patterns in a case expression is assumed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

@@

%\vspace{-0.5cm}
\[
\TypeVarDelta
\]
The rule states that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is made up exactly of the resources in the usage environment of
that variable, \ROUNDTWO{denoted by the $\Delta = \Delta'$ premise of
  the rule.}
%
\ROUNDTWO{$\Delta$-bound variables} are unrestricted and can be discarded and duplicated, despite
multiple occurrences of the \ROUNDTWO{variable} in an expression not
being well-typed
\ROUNDTWO{due to} non-linear usage of linear resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

@@

In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
resources used in let-bound expressions are only consumed when the respective
binders are evaluated in the let body.
%, i.e. linear resources required by let-bound expressions are consumed lazily.
%
Moreover, resources captured by a let-bound expression cannot be used
simultaneously with the let-binder in its body, since \ROUNDTWO{those}
resources would end up being consumed more than once and violate
\ROUNDTWO{the semantics of}
linearity. % The binder has to be used in mutual exclusion with the linear
% resources captured by the expression it binds --
Either the binding or the
resources \emph{must} be used.

% Indeed, usage environments allow us to encode mutual exclusivity between
% alternative ways of consuming linear resources (between $\D$-vars and direct
% resource usage). 
Let-bound variables are the canonical example of a
$\Delta$-bound variable.
% , that is, let-bound variables bind expressions in which
% the resources required to type them are consumed lazily rather than eagerly.
%
Annotating let-bound variables with \ROUNDTWO{their} usage environment $\D$ delays the
consumption of resources to when the variables themselves are used and
guarantees the desired mutual exclusion property between the binder and the resources:
\[
\TypeLet
\]
The rule for (non-recursive) let bindings splits the linear environment in $\D$
and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
surprisingly, the resources $\D$ used to type $e$ are still available in the
environment to type the let body $e'$, alongside the unrestricted $x$ binding
annotated with the usage environment $\D$. In essence, the resources of $e$ being
available in $e'$ reflects how typing the body of a lazy binder
doesn't consume resources, and the binding $x$ being an unrestricted
$\D$-variable ensures that using it will consume the resources $\D$ the
expression captures, \emph{if} it is used.
% \todo[inline]{Let bindings are hard, if they are used then we use resources. If
% they don't get used then we use no resources! In practice, resources that show
% up in the body of the let must be used, be it by using the let binder, or by
% using them directly. This makes the let binder and the resources in its body
% mutual exclusive.}
